The role of advection versus diffusion in intraventricular transport

### 2025-02-17 paper/main.tex
add r/lambda table to appendix

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from 27.7 \textmu m/s to 8.0 \textmu m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Changes in the transport dynamics of Dendra2 are most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes in the dynamics of $\cbar$ (\Cref{fig:fig5}b).
Compared to the baseline model results, we observe a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia does not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:fig5}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions where loss of cilia motility is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    loss-of-cilia-motility scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

@@

The tangential stresses $\btau = \tau\lambda(\xx) P_{\nn}(\rr)$ that represent the cilia
motion differ for the cilia regions. We heuristically determined $\tau$ and $\lambda(\xx)$.
As noted previously, $\tau=\num{6.5e-4}$ Pa. We set $\rr = \rr^+$ and $\rr = \rr^-$ in
the ventral and dorsal regions of the diencephalic ventricle, respectively. In the ventral region,
we set the linear profile $\lambda(\xx) = \frac{1}{2}\left(1 - \frac{x-x_{0, v}}{x_{e, v} - x_{0, v}}\right)$,
where $[x_{0, v}, x_{e, v}]$ is the $x$-range of the cilia region, so the stresses decrease
linearly towards the posterior part of the geometry. We impose a linearly increasing profile
over the $x$-range $[x_{0, d}, x_{e, d}]$ in the dorsal region:
$\lambda(\xx)=\frac{11}{4}\left(\frac{x-x_{0, d}}{x_{e, d} - x_{0, d}}\right)$.
\lyng{Put all of this information on $\lambda$ choice in a table.}

In the telencephalic ventricle, we set $\rr=\rr^-$ and $\lambda=\frac{2}{5}$
in the anterior most part of the cilia region,
and $\rr=\rr^+$ and $\lambda(\xx)=1$ in the posterior part.

@@

\subsection{Numerical verification of model implementation}\label{subsec:numerical_verification}
A cylinder geometry meshed with varying degrees of refinement was used to
perform convergence analysis of the numerical methods used to solve the Stokes equations.
The cylinder had unit length and a diameter of one half length unit.
For the analysis, we used tangential traction boundary conditions on all
of the cylinder surface. Using the Method of Manufactured Solutions~\cite{Roache2001CodeSolutions}
with the velocity and pressure expressions defined in~\Cref{subsec:mms}
for $\uu_{\mathrm{ex}}$ and $p_{\mathrm{ex}}$, we calculated $\mathbf{H}(\mathrm{div};\Omega)$
and $\mathbf{H}^1$ error norms for $\uu_h$, and the $L^2$ error norm for
$p_h$~(\Cref{tab:cylinder_error_rates}). Based on the error norm calculations,
we also calculated the experimental order of convergence~(\Cref{subsec:norms_and_eoc}).
We observe convergence in all of the reported error norms. Furthermore,
we report the error in calculation of the maximum velocities, and observe
convergence in this error as well. Lastly, we calculated the $L^2$ norm of
the divergence $\nabla\cdot\uu_h$ to verify the property of mass conservation.
\begin{table}[!htbp]
    \small
    \centering
    \caption{Error norms of the velocity $\uu_h$ in the
    $\mathbf{L}^2$ and $\mathbf{H}^1$ norms,
    the pressure $p_h$ in the $L^2$ norm,
    and the error in the maximum velocity magnitude $\uu_{h, {\mathrm{mag}}}$
    for mesh refinement on a cylinder mesh with the only-cilia/no-cardiac flow model.
    Additionally, we report the $L^2$ norm of the divergence of the velocity.
    For all error norms, convergence rates calculated
    with~\eqref{eq:convergence_order_estimate} are given
    in parentheses.}\label{tab:cylinder_error_rates}
    \begin{tabular}{cc|ccccc}
        \toprule
        $h/h_0$ & \# dofs &
        $\normltwovec{\uu_h - \uu_{\mathrm{ex}}}$
        & $\Hnorm{\uu_h - \uu_{\mathrm{ex}}}$
        & $\normltwo{p_h - p_{\mathrm{ex}}}$
        & $\normlinf{u_h} -\normlinf{u_{\mathrm{ex}}}$
        & $\normltwo{\nabla\cdot\uu_h}$\\ 
        \midrule 
        1    & 8652    & \num{2.3e-02}  (--)  & \num{4.3e-01}   (--)  & \num{5.4e-01}  (--) 
        & \num{2.4e-02} (---) & \num{1.4e-07} \\

>>>>

The CSF velocities are most strongly
affected when removing cilia in the dorsal diencephalic ventricle,
as the maximum velocity magnitude drops from 27.7 \textmu m/s to 8.0 \textmu m/s.
\emph{In vivo}, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing telencephalic and ventral diencephalic cilia has minimal
impact and results in maximum speeds of 27.7 \textmu m/s or 27.2 \textmu m/s, respectively.
Changes in the transport dynamics of Dendra2 are most prominent when removing the dorsal cilia
in the diencephalic ventricle. All ROIs (except ROI 1, where we impose the photoconversion curve)
exhibit changes in the dynamics of $\cbar$ (\Cref{fig:fig5}b).
Compared to the baseline model results, we observe a slower increase and lower final concentration
in the middle and anterior ventricles (ROIs 2--5),
and more rapid transport to the posterior ventricle (ROI 6).
In contrast, removal of telencephalic cilia does not alter the evolution of the
mean concentration $\cbar$ with time (\Cref{fig:fig5}b).
Finally, without ventral cilia in the diencephalic ventricle,
more solute spreads towards the anterior ventricle (ROI 5) and
less solute spreads towards the posterior ventricle (ROI 6).
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure5_compare_cilia_modifications.png}
    \caption{\textbf{a.} The three separate ciliated regions cilia paralysis is simulated
    by imposing a zero tangential traction boundary condition in the respective region.
    \textbf{b.} Mean concentration $\cbar(t)$ for the six regions of interest (ROIs) in the three 
    paralysis scenarios considered. The region named in the legend is the region
    where we modify the cilia tangential traction boundary condition to a free-slip condition
    in the CSF flow model.}
    \label{fig:fig5}
\end{figure}

@@

The tangential stresses $\btau = \tau\lambda(\xx) P_{\nn}(\rr)$ that represent the cilia
motion differ for the cilia regions. We heuristically determined the constant $\tau= 0.65$ mPa
and the spatially varying function $\lambda(\xx)$. The definitions of $\rr$ and $\lambda(\xx)$
for the different cilia populations are summarized in~\Cref{tab:r_vector_and_lambda_function}.
\begingroup
\setlength{\tabcolsep}{20pt}
\renewcommand{\arraystretch}{1.5}
\begin{table}[H]
    \centering
    \caption{Definitions of $\rr$ and $\lambda(\xx)$ used in for the cilia tangential 
    traction boundary condition $\btau = \tau\lambda(\xx)P_{\nn}(\rr)$,
    where $\tau=0.65$ mPa. The $x-$coordinates $x_{0, i}$ and $x_{e, i}$ define 
    the cilia domain in the dorsal ($i=d$) and ventral ($i=v$) regions of 
    the middle ventricle.}\label{tab:r_vector_and_lambda_function}
    \begin{tabular}{lcc}
        \toprule
        Cilia population & $\rr$ & $\lambda(\xx)$ \\
        \midrule 
        Anterior region, anterior ventricle  & $-(1, 0, 1)$ & $\frac{2}{5}$ \\
        Posterior region, anterior ventricle & $\ \ \ (1, 0, 1)$ & 1 \\
        Dorsal region, middle ventricle      & $-(1, 0, 1)$ & $\lambda(\xx)=\frac{11}{4}\left(\frac{x-x_{0, d}}{x_{e, d} - x_{0, d}}\right)$\\
        Ventral region, middle ventricle     & $\ \ \ (1, 0, 1)$ & $\lambda(\xx) = \frac{1}{2}\left(1 - \frac{x-x_{0, v}}{x_{e, v} - x_{0, v}}\right)$ \\
        \bottomrule
    \end{tabular}
\end{table}
\endgroup

@@

\subsection{Numerical verification of model implementation}\label{subsec:numerical_verification}
A cylinder geometry meshed with varying degrees of refinement was used to
perform convergence analysis of the numerical methods used to solve the Stokes equations.
The cylinder had unit length and a diameter of one half length unit.
For the analysis, we used tangential traction boundary conditions on all
of the cylinder surface. Using the Method of Manufactured Solutions~\cite{Roache2001CodeSolutions}
with the velocity and pressure expressions defined in~\Cref{subsec:mms}
for $\uu_{\mathrm{ex}}$ and $p_{\mathrm{ex}}$, we calculated $\mathbf{L}^2$
and $\mathbf{H}^1$ error norms for $\uu_h$, and the $L^2$ error norm for
$p_h$~(\Cref{tab:cylinder_error_rates}). Based on the error norm calculations,
we also calculated the experimental order of convergence~(\Cref{subsec:norms_and_eoc}).
We observe convergence in all of the reported error norms. Furthermore,
we report the error in calculation of the maximum velocities, and observe
convergence in this error as well. Lastly, we calculated the $L^2$ norm of
the divergence $\nabla\cdot\uu_h$ to verify the property of mass conservation.
\begin{table}[!htbp]
    \small
    \centering
    \caption{Error norms of the velocity $\uu_h$ in the
    $\mathbf{L}^2$ and $\mathbf{H}^1$ norms,
    the pressure $p_h$ in the $L^2$ norm,
    and the error in the maximum velocity magnitude $\uu_{h, {\mathrm{mag}}}$
    for mesh refinement on a cylinder mesh with the only-cilia/no-cardiac flow model.
    Additionally, we report the $L^2$ norm of the divergence of the velocity.
    For all error norms, convergence rates calculated
    with~\eqref{eq:convergence_order_estimate} are given
    in parentheses.}\label{tab:cylinder_error_rates}
    \begin{tabular}{cc|ccccc}
        \toprule
        $h/h_0$ & \# dofs &
        $\normltwovec{\uu_h - \uu_{\mathrm{ex}}}$
        & $\Hnorm{\uu_h - \uu_{\mathrm{ex}}}$
        & $\normltwo{p_h - p_{\mathrm{ex}}}$
        & $\normlinf{u_h} -\normlinf{u_{\mathrm{ex}}}$
        & $\normltwo{\nabla\cdot\uu_h}$\\ 
        \midrule 
        1    & 8652    & \num{2.3e-02}  (--)  & \num{4.3e-01}   (--)  & \num{5.4e-01}  (--) 
        & \num{2.4e-02} (---) & \num{1.4e-07} \\
